---
title: "STA 561 HW2 (Decision Trees)"
author: "Daniel Truver"
date: "1/29/2018"
header-includes:
  - \usepackage{amsmath}
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

#### (1) Classifiers for Basketball Courts  

```{r Perceptron}
perceptron = function(X, y, I){
  iteration = 1
  w = rep(0, ncol(X))
  for (j in 1:I){
    for (i in seq_along(y)){
      if (y[i]*(w %*% X[i,]) <= 0){
        w = w + y[i]*X[i,] # update step
        iteration = iteration + 1
      }
    }
    accuracy = sum((X %*% w) * y > 0)/length(y) # calculate propotion of correctly classified point
    if (accuracy == 1){ 
      break # no need to continue if we have perfect separation
    } 
  }
  return(list("iteration" = iteration, "w" = w, "accuracy" = accuracy))
}
```

##### (a) Let's run the perceptron because it's fun. 

```{r enteringB-BallData}
X_1 = c(.75, .85, .85, .15,.05,.05,.85)
X_2 = c(.1,.8,.95,.1,.25,.5,.25)
Y = c(-1, -1, 1, -1, 1, 1, -1)
X_b = cbind(X_1, X_2)
```

```{r perceptron}
res = perceptron(X_b, Y, I = 100)
```

The preceptron made `r res$iteration` mistakes before converging with accuracy = `r res$accuracy` (error = `r 1-res$accuracy`). See the decision boundary below. 

```{r perceptronDecisionBound}
suppressMessages(library(ggplot2))
ggplot(data = data.frame(X_1, X_2, Y), aes(x = X_1, y = X_2)) +
  geom_point(aes(pch = factor(Y)), size = 7) +
  scale_shape_manual(values = c("-", "+")) +
  geom_abline(slope = -res$w[["X_1"]]/res$w[["X_2"]], intercept = 0) +
  ggtitle("Result of Perceptron Algorithm",
          subtitle = "Other possible separators featured in color") +
  geom_abline(intercept = 0, slope = 1, color = "red", lty = "dashed") + 
  geom_abline(intercept = 0, slope = 1.05, color = "blue", lty = "dashed") + 
  theme_bw() + 
  theme(plot.subtitle = element_text(color = "red"))
```  

##### (b) Growing the decision tree.

```{r testGrow}
gini.index = function(node){ # function to calculate gini Index
  p = sum(node > 0)/length(node)
  I = 2*p*(1-p)
  return(I)
}
X = X_b
Y = Y
grow.tree = function(X, Y, current_node = NULL, 
                     epsilon = 10^-6, threshold = 0.20){ # split node based on gini index
  X = data.frame(X)
  Y = data.frame(Y = Y)
  data.df = cbind(X,Y)
  if (is.null(current_node)){
    current_node = data.df
  }
  # variables to be filled later
  delta.I.record = c()
  leaf1.record = list()
  leaf2.record = list()
  split.record = list()
  c = 1
  for ( i in 1:(ncol(current_node)-1) ){
    # get possible split values, adjust min, max up, down to make inclusive inequalities 
    # run without trouble
    split.sequence = sort(unique(current_node[,i]))
    split.sequence[1] = split.sequence[1] + epsilon
    split.sequence[length(split.sequence)] = split.sequence[length(split.sequence)] - 
      epsilon
    for(x in split.sequence){ # this loop calculates each possible split and reduction in gini index
      split.record = append(split.record, list(c(i, x)))
      node_1 = current_node[which(current_node[,i] <= x),]
      leaf1.record[[c]] = node_1
      I_1 = gini.index(node = node_1$Y)
      node_2 = current_node[which(current_node[,i] > x),]
      leaf2.record[[c]] = node_2
      I_2 = gini.index(node = node_2$Y)
      delta.I = gini.index(current_node$Y) - sum(nrow(node_1)/nrow(current_node)*I_1,
                                                 nrow(node_2)/nrow(current_node)*I_2)
      delta.I.record = c(delta.I.record, delta.I)
      c = c + 1
    }
  }
  optimal = which.max(delta.I.record) # find split with greatest gini reduction
  new.nodes = list(leaf1.record[[optimal]], leaf2.record[[optimal]])
  Split = split.record[[optimal]]
  return(list("opt" = delta.I.record[optimal], "leaves" = new.nodes, "split" = Split,
              "threshold" = delta.I.record[optimal] > threshold))
}
```

The function defined above splits a single node based on the gini index. We implement it below, step by step, until we have grown a full tree.

```{r growingFullTree}
tree_1 = grow.tree(X,Y)
pure_1 = length(unique(tree_1$leaves[[1]]$y)) == 1
pure_2 = length(unique(tree_1$leaves[[2]]$y)) == 1
split_1 = tree_1$split
# we see that the second leaf is not pure, so we attempt to split it again
tree_2 = grow.tree(X, Y, current_node = tree_1$leaves[[2]])
pure_1 = length(unique(tree_2$leaves[[1]]$y)) == 1
pure_2 = length(unique(tree_2$leaves[[2]]$y)) == 1
split_2 = tree_2$split
# all nodes are now pure; we should stop
```

We can see the splits in this lovely graphic. 

```{r plottingTheTree, echo=FALSE}
ggplot(data = data.frame(X_1, X_2, Y), aes(x = X_1, y = X_2)) +
  geom_point(aes(pch = factor(Y)), size = 7) +
  scale_shape_manual(values = c("-", "+")) +
  geom_vline(xintercept = tree_1$split[2], lty = "dotted") +
  geom_text(aes(x = tree_1$split[2]+.05, y = 0.7, label = "X_1 <= .05")) +
  geom_line(data=data.frame(x = c(tree_1$split[2],1),y = tree_2$split[2]), aes(x,y), lty = "dotted") +
  geom_text(aes(x = 0.5, y = tree_2$split[2]+.05, label = "X_2 > 0.80")) +
  theme_bw() +
  ggtitle("Decision Tree Splits on Basketball Court Data",
          subtitle = "Equal Accuracy Decision Tree (for this dataset) in red") +
  geom_vline(xintercept = 0.075, color = "red", lty = "dotted") +
  geom_line(data = data.frame(x = c(.075,1),y = .825), aes(x,y), color = "red", lty = "dotted") +
  theme(plot.subtitle = element_text(color = "red"))
```

The error on this tree is 0. We should note that if we set the $\Delta I$ threshold at or above $0.27$, then there are not splits in the tree; we get only one class. 

##### (c) The optimal linear classfier

If we want to minimize the expected loss, we need to minimize the area between the true boundary, $\sqrt{x_1}$ and the decision boundary $w\cdot x_1$. Therefore, we compute the following integral.  

$$
\begin{aligned}
h(w) 
&= \int_0^1\min(|\sqrt{x_1} - wx|, 1-\sqrt{x_1})dx_1 \\
&= \int_0^{1/w^2} (\sqrt{x_1}-wx_1)dx_1 + \int_{1/w^2}^{1/w}(wx_1-\sqrt{x_1})dx_1 + \int_{1/w}^1(1-\sqrt{x_1})dx_1 \\
&= \frac{1}{3}w^{-3} - \frac{1}{2}w^{-1} + \frac{1}{3}
\end{aligned}
$$

To get the minimum, we take the derivative, set that baby equal to zero, and crank the algebra machine until we get: 

$$
\begin{aligned}
& 0 = h'(w) = -w^{-4} + \frac{1}{2}w^{-2} \\
&\implies w = 2^{1/2} = 1.414
\end{aligned}
$$

We plot the optimal decision boundary here. 

```{r plottingOptimal, echo=FALSE}
x_1.plot = seq(0,1, length.out = 100)
x_2.plot = sqrt(x_1.plot)
ggplot(data = data.frame(X_1, X_2, Y), aes(x = X_1, y = X_2)) +
  geom_point(aes(pch = factor(Y)), size = 7) +
  scale_shape_manual(values = c("-", "+")) +
  geom_abline(slope = 1.414, intercept = 0) + 
  geom_line(data = data.frame(x_1.plot, x_2.plot), aes(x = x_1.plot, y = x_2.plot),
            lty = "dotted")+
  geom_abline(slope = -res$w[["X_1"]]/res$w[["X_2"]], intercept = 0, 
              color = "red", lty = "dashed") +
  theme_bw() +
  ggtitle("Optimal Linear Classifier",
          subtitle = "Our perceptron decision boundary in red") +
  theme(plot.subtitle = element_text(color = "red")) +
  xlim(0,1) +
  ylim(0,1)
```

We note that this classifier has lower accuracy in than the perceptron separator we found in a previous part. This is not surprising; the optimal classifier is the one that should perform best on average over a set of datasets, whereas our perceptron trained on one dataset specifically.  

##### (d) The optimal decision tree

The general idea behind our approach here is the same as in optimizing the linear classifier. Since we are assuming the location of shots to be uniformly distributed, we want to minimize the misclassified area of the court. We begin with a visulalization. A general decision tree that splits on $X_1$ and then $X_2$ looks something like this. 

```{r generalDecisionTree, echo=FALSE}
#points for tree boundaries 
S1 = data.frame(x_1 = c(.5,.5), x_2 = c(0,1))
S2 = data.frame(x_1 = c(0,.5), x_2 = c(.5,.5))
S3 = data.frame(x_1 = c(.5,1), x_2 = c(.875,.875))
ggplot(data = data.frame(x_1.plot, x_2.plot), aes(x = x_1.plot, y = x_2.plot)) +
  geom_line() +
  geom_line(data = S1, aes(x_1, x_2), lty = "dashed") +
  geom_line(data = S2, aes(x_1, x_2), lty = "dashed") +
  geom_line(data = S3, aes(x_1, x_2), lty = "dashed") +
  geom_text(data = data.frame(x = c(.25,.25,.75,.75), y = c(.25,.75,.5,.95)),
            aes(x = x, y = y, label = c("(-)","(+)","(-)","(+)")),
            size = 14, color = "blue") +
  geom_text(data = data.frame(x = c(.05,.42,.57,.95), y =c(.375,.55,.8,.92)),
            aes(x=x, y =y, label = "PROBLEM")) +
  geom_text(aes(x = -.02, y = .5, label = "s_2")) +
  geom_text(aes(x = .5, y = 0, label = "s_1")) +
  geom_text(aes(x = 1.02, y = .875, label = "s_3")) +
  theme_bw() +
  ggtitle("General flavor of decision tree split on X_1 first")
```

Here, $s_1$ is the $x_1$ intercept of the vertical split, and $s_2, s_3$ are the $x_2$ intercepts of the left and right horizontal splits, respectively. Classification of each region is given by the obnoxiously large positive and negative signs. The ares that would be misclassified have a large 'PROBLEM' label. 

We can calculate this area as follows:
$$
\begin{aligned}
h(s_1, s_2, s_3) 
&= \int_0^{s_1} |\sqrt{x_1} - s_2|dx_1 + \int_{s_1}^1 |\sqrt{x_1}-s_3|dx_1 \\
&= \int_0^{s_2^2}(s_2 - \sqrt{x_1})dx_1 + \int_{s_2^2}^{s_1}(\sqrt{x_1}-s_2)dx_1 \\
&\quad\quad +\int_{s_1}^{s_3^2}(s_3 - \sqrt{x_1})dx_1 + \int_{s_3^2}^1(\sqrt{x_1}-s_3)dx_1 \\
&\text{...some algebra later...} \\
&= \frac{2}{3}s_2^3 + \frac{2}{3}s_3^3 + \frac{4}{3}s_1^{3/2} - s_2s_1 -s_3s_1 -s_3 + \frac{2}{3}
\end{aligned}
$$

Now, we could take some partial derivatives, set things equal to zero, do more algebra, pray for death, and get an honest to (your preferred deity here) expression for these variables. Or, in less than one second, we run the following code chunk and get an approximation.

```{r exhaustiveMinimization}
s1 = seq(0,1, length.out = 100)
s2 = seq(0,1, length.out = 100)
s3 = seq(0,1, length.out = 100)
split.exhaustive = matrix(NA, nrow = 100^3, ncol = 4)
i = 1
for (x in s1){
  for (y in s2){
    for (z in s3){
      split.exhaustive[i, 1] = x
      split.exhaustive[i, 2] = y
      split.exhaustive[i, 3] = z
      split.exhaustive[i, 4] = (2/3)*y^3 + (2/3)*z^3 + (4/3)*x^(3/2) -y*x -z*x - z + 2/3
      i = i + 1
    }
  }
}
minimum = which.min(split.exhaustive[,4])
s.opt = split.exhaustive[minimum, 1:3]
```

```{r tableForSopt, echo=FALSE}
rm(split.exhaustive)
knitr::kable(data.frame(s_1 = s.opt[1], s_2 = s.opt[2], s_3 = s.opt[3]),
             caption = "Optimal Splitting Values", digits = 2)
```  

Please enjoy the following graphic. Signed copies available upon request.

```{r optimalTree1Visual, echo=FALSE}
ggplot(data = data.frame(X_1, X_2, Y), aes(x = X_1, y = X_2)) +
  geom_point(aes(pch = factor(Y)), size = 7) +
  scale_shape_manual(values = c("-","+")) +
  geom_vline(xintercept = s.opt[1]) +
  geom_line(data = data.frame(x = c(0,s.opt[1]), y = s.opt[2]), aes(x,y)) +
  geom_line(data = data.frame(x = c(s.opt[1], 1), y = s.opt[3]), aes(x,y)) +
  ggtitle("Theoretical Optimal Decision Tree",
          subtitle = "Our grown decision tree in red") +
  geom_vline(xintercept = tree_1$split[2], color = "red", lty = "dashed") +
  geom_line(data=data.frame(x = c(tree_1$split[2],1),y = tree_2$split[2]), 
            aes(x,y), lty = "dashed", color = "red") +
  geom_line(data = data.frame(x_1.plot, x_2.plot), aes(x_1.plot, x_2.plot),
            lty = "dotted") +
  theme_bw() +
  theme(plot.subtitle = element_text(color = "red")) 
```

Once again, we see that the accuracy is higher on our grown tree for this particular dataset. The theoretically optimal tree only has lower expected loss on average.  

Now, we move on to the case of splitting on $X_2$ first. The general case looks something like this:

```{r generalTree2, echo=FALSE}
ggplot(data=data.frame(x_1.plot, x_2.plot), aes(x_1.plot, x_2.plot)) +
  geom_line(lty = "dotted") +
  geom_hline(yintercept = .5) +
  geom_line(data = data.frame(x = .125, y = c(0,.5)), aes(x,y)) +
  geom_line(data = data.frame(x = .6, y = c(.5,1)), aes(x,y)) +
  geom_text(data = data.frame(x = c(.05,.1, .8,.8), y = c(.25,.75,.25,.75)),
            aes(x,y, label = c("(+)","(+)", "(-)","(-)")), color = "blue", size = 10) +
  geom_text(data = data.frame(x = c(0, .125, .6), y = c(.55, -.02, 1.02)), 
            aes(x,y, label = c("s_1", "s_2", "s_3"))) +
  ggtitle("General flavor of decision tree split on X_2 first") +
  theme_bw()
```  

We will go through the same process as before to obtain an expression for the expected loss. 
$$
\begin{aligned}
h(s_1, s_2, s_3)
&= \int_0^{s_2} \sqrt{x_1}dx_1 + \int_{s_2}^{s_1^2}(s_1 - \sqrt{x_1})dx_1 + \int_{s_1}^{s_3}(\sqrt{x_1} - s_1) dx_1+ \int_{s_3}^1(1-\sqrt{x_1})dx_1\\
&\text{...some algegra later...}\\
&= \frac{4}{3}s_2^{3/2} + \frac{4}{3}s_3^{3/2} + \frac{1}{3}s_1^3 - s_1s_2 -s_1s_3-s_3 + \frac{1}{3}
\end{aligned}
$$

Again, we will compute the minimum of this function with this lovely computing machine exactly as above to obtain the following values for $s_1, s_2, s_3$. 

```{r exhaustiveTree2, echo=FALSE}
split.exhaustive = matrix(NA, nrow = 100^3, ncol = 4)
i = 1
for (x in s1){
  for (y in s2){
    for (z in s3){
      split.exhaustive[i, 1] = x
      split.exhaustive[i, 2] = y
      split.exhaustive[i, 3] = z
      split.exhaustive[i, 4] = (4/3)*y^(3/2) + (4/3)*z^(3/2) + (2/3)*x^(3) -y*x -z*x - z + 1/3
      i = i + 1
    }
  }
}
minimum = which.min(split.exhaustive[,4])
s.opt = split.exhaustive[minimum, 1:3]
knitr::kable(data.frame(s_1 = s.opt[1], s_2 = s.opt[2], s_3 = s.opt[3]),
             caption = "Optimal values for tree split on X_2 first", digits = 2)
```

Visually, it looks something like this.

```{r plottingOptTree2, echo=FALSE}
ggplot(data = data.frame(X_1, X_2, Y), aes(x = X_1, y = X_2)) +
  geom_point(aes(pch = factor(Y)), size = 7) +
  scale_shape_manual(values = c("-","+")) +
  geom_hline(yintercept = s.opt[1]) +
  geom_line(data = data.frame(x = s.opt[2], y = c(0,s.opt[1])), aes(x,y)) +
  geom_line(data = data.frame(x = s.opt[3], y = c(s.opt[1], 1)), aes(x,y)) +
  geom_line(data = data.frame(x_1.plot, x_2.plot), aes(x_1.plot, x_2.plot),
            lty = "dotted") +
  ggtitle("Theoretical Optimal Decision Tree",
          subtitle = "Our grown decision tree in red") +
  geom_vline(xintercept = tree_1$split[2], color = "red", lty = "dashed") +
  geom_line(data=data.frame(x = c(tree_1$split[2],1),y = tree_2$split[2]), 
            aes(x,y), lty = "dashed", color = "red") +
  geom_line(data = data.frame(x_1.plot, x_2.plot), aes(x_1.plot, x_2.plot),
            lty = "dotted") +
  theme_bw() +
  theme(plot.subtitle = element_text(color = "red")) 
``` 

Once again, the accuracy is lower for the optimal loss tree on this specific example. We are still not troubled by this for the reasons explained above. 

NOTE: In the above integrals, we have assumed each time that $s_2 < s_3$. We are comfortable making this decision since moving the $s_2$ and $s_3$ lines past each other always results in more misclassified area. Namely, an arrangement such as this will have the $s_2$ or $s_3$ line intersect the $s_1$ line without intersecting the true decision boundary. This will yield an additional rectangle of misclassified area that need not exist.  

##### (e) Transformations of Covariates  

Well, this one seems fairly tame after the algebra monstrosity of part (d). We know the true decision boundary to be $x_2 = \sqrt{x_1}$, so $x_2^2 = x_1$, and we consider the transformation, $t:x_2 \rightarrow x_2^2$. 

```{r transformPlot, echo=FALSE}
ggplot(data = data.frame(X_1, X_2 = X_2^2, Y), aes(X_1, X_2)) +
  geom_point(aes(pch = factor(Y)), size = 7) +
  scale_shape_manual(values = c("-", "+")) +
  geom_line(data = data.frame(x_1 = x_1.plot, x_2 = x_2.plot^2), aes(x_1, x_2)) +
  geom_abline(slope = 1, intercept = 0, color = "red", lty = "dashed", 
              size = 1.2, alpha = .8) +
  ggtitle("Transformation of x_2 to x_2^2",
          subtitle = "Perfect Classifier in red") +
  theme_bw() +
  theme(plot.subtitle = element_text(color = "red")) +
  xlab("t(X_2)")
```

Using the recognizing straight lines property, we deduce that the optimal classifier is the line of slope 1 passing through the origin under the transformation $t:x_2 \rightarrow x_2^2$. This classifier will always have zero error (perfect accuracy) because it is the same as the decision boundary.

##### (f) Decision tree under transformation

We could not obtain the same error under this transformation, especially not a depth 2 decision tree. A tree will produce a finite number of rectangular partitions of the space which are unable to capture a diagonal line. One could attempt an argument for 0 expected loss as the number of terminal nodes $c \to\infty$, but this is unrealistic.

##### (h) Classifying the paint (linear)

We will use the same integration technique as above. We consider below the line as inside the paint and above the line as outside the paint. We want to minimize the area
$$
\begin{aligned}
h(w) 
&= \int_0^{.5} wx_1 dx_1 + \int_{.5}^{.25/w} (.25-wx_1)dx_1 + \int_{.25/w}^{1} (wx_1 - .25)dx_1 \\
&\text{...some algebra later...}\\
&= \frac{1}{16}w^{-1} + \frac{1}{4}w + \frac{3}{8}
\end{aligned}
$$

Setting $h'(w) = 0$, we get 
$$
w = \sqrt{\frac{1}{12}}
$$

Which looks like

```{r paintLlinearPlot, echo=FALSE}
trueRisk = (1/16)/(sqrt(1/12)) + (3/4)*sqrt(1/12) - 3/8
ggplot(data = data.frame(x = c(.5,.5,1), y =c(0,.25,.25) ), aes(x,y)) +
  geom_line(lty = "dotted") +
  xlim(0,1) + 
  ylim(0,1) +
  geom_abline(intercept = 0, slope = sqrt(1/12)) +
  theme_bw() +
  ggtitle("Optimal Linear Classifier for the Paint",
          "Below line classified as +1") +
  ylab("x_2") + xlab("x_1")
``` 

The 'true risk' of this classfier is `r round(trueRisk, 2)`.

##### (i) Classifying the paint (tree)

Fortunately, the structure of the decision tree decision boundary makes this obvious.  

$$
f(x_1, x_2) = 
\begin{cases}
-1 & x_1\leq 0.5 \\
-1 & (x_1, x_2) \in (0.5,1] \times (0.25, 1]  \\
1 & (x_1, x_2) \in (0.5,1] \times [0, 0.25]
\end{cases}
$$

The decision boundary is as follows. 

```{r paintTreePlot}
ggplot(data = data.frame(x = c(.5,.5,1), y =c(0,.25,.25) ), aes(x,y)) +
  geom_line() +
  xlim(0,1) + 
  ylim(0,1) + 
  ggtitle("Paint Decision Boundary", 
          "Optimal Tree Decision Boundary in Red") +
  geom_vline(xintercept = 0.5, color = "red", lty = "dashed", size = 1, alpha = .65) +
  geom_line(data = data.frame(x_1 = c(0.5, 1), x_2 = .25), aes(x_1, x_2),
            color = "red", lty = "dashed", size = 1, alpha = .65) +
  theme_bw() +
  theme(plot.subtitle = element_text(color = "red")) +
  geom_text(aes(.75, .125, label = "(+1)"), size = 6, color = "red") +
  geom_text(aes(.75, .75, label = "(-1)"), size = 6, color = "red") +
  geom_text(aes(.25, .5, label = "(-1)"), size = 6, color = "red") +
  ylab("x_2") + xlab("x_1")
```  

The error is clearly 0.

******

#### (2) Variable Importance for Trees and Forests  

Hurrah! CSV files!

```{r readInData}
train = read.csv("train.csv")
test = read.csv("test.csv")
```

##### (a) A Single Stump  

Fine, I'll write my own. It's going to be specific to this prblem though. I'll work on the generality later. The Gini Index function here will be different that question (1). It will accept data frames rather than a vector for the sake of avoiding syntax problems and maintaining sanity. 

```{r growStump}
gini.index = function(node.df){ # response should be called Y, always {0,1}
  p = sum(node.df$Y)/nrow(node.df)
  I = 2*p*(1-p)
  return(I)
}
grow.stump = function(df, surrogate = TRUE){ # df = data frame of response AND covariates
  N = nrow(df)
  I_df = gini.index(df) # get the gini index value of the full data
  predictors = names(df)[names(df) != "Y"]
  c = 1
  leaf1.record = NULL # new strategy for storing results 
  leaf2.record = NULL 
  delta.I.record = -Inf  
  for (j in predictors){ # no need for inner loop since all covariates are binary
    predictor.is.1 = which(df[,j] == 1)
    predictor.is.0 = -predictor.is.1
    leaf1 = df[predictor.is.0,]
    leaf2 = df[predictor.is.1,]
    class1 = 1*(sum(leaf1$Y > 0) > sum(leaf1$Y <= 0))
    class2 = 1*(sum(leaf2$Y > 0) > sum(leaf2$Y <= 0))
    I_1 = gini.index(leaf1)
    I_2 = gini.index(leaf2)
    N_1 = nrow(leaf1)
    N_2 = nrow(leaf2)
    delta.I = I_df - sum( (N_1/N)*I_1, (N_2/N)*I_2 )
    if (delta.I > delta.I.record){
      delta.I.record = delta.I
      leaf1.record = list("leaf" = leaf1, "split" = j, "class" = class1)
      leaf2.record = list("leaf" = leaf2, "split" = j, "class" = class2)
    }
  }
  surrogateStump = NULL
  if (surrogate){
    j = actualBest = leaf1.record$split
    surrogateOptions = predictors[!(predictors %in% actualBest)]
    lambda = rep(NA, length(surrogateOptions))
    l = 1
    for (j.tilde in surrogateOptions){
      p_L = nrow(leaf1.record$leaf)/N
      p_R = nrow(leaf1.record$leaf)/N
      p_min = min(p_L, p_R)
      p_LL = sum(df[,j] == 0 & df[,j.tilde] == 0)/N
      p_RR = sum(df[,j] == 1 & df[,j.tilde] == 1)/N
      p_minmin = 1 - p_LL - p_RR
      lambda[l] = (p_min - p_minmin)/p_min
      l = l+1
    }
    bestSurrogate = surrogateOptions[which.max(lambda)]
    surrogateSplit.1 = which(df[,bestSurrogate] == 1)
    surrogateSplit.0 = -surrogateSplit.1
    surrogateLeaf1 = df[surrogateSplit.0,]
    surrogateLeaf2 = df[surrogateSplit.1,]
    class1 = 1*(sum(surrogateLeaf1$Y > 0) > sum(surrogateLeaf1$Y <= 0))
    class2 = 1*(sum(surrogateLeaf2$Y > 0) > sum(surrogateLeaf2$Y <= 0))
    deltaSurrogate = I_df - sum(nrow(surrogateLeaf1)/N*gini.index(surrogateLeaf1),
                                nrow(surrogateLeaf2)/N*gini.index(surrogateLeaf2))
    surrogateStump = list("leaf1" = surrogateLeaf1, 
                          "leaf2" = surrogateLeaf2,
                          "split" = bestSurrogate,
                          "class1" = class1, "class2" = class2,
                          "delta" = deltaSurrogate)
  }
  return(list("leaf1" = leaf1.record, "leaf2" = leaf2.record, "deltaI" = delta.I.record,
              "surrogate" = surrogateStump))
}
```

##### (a)(i) Best and Surrogate Splits

```{r firstStumps}
stump1 = grow.stump(train)
cat("Best split:", stump1$leaf1$split, 
    "\n with '0' leaf predicting:", stump1$leaf1$class,
    "\n and '1' leaf predicting:", stump1$leaf2$class)
cat("\nSurrogate Split:", stump1$surrogate$split,
    "\n with '0' leaf predicting:", stump1$surrogate$class1,
    "\n and '1' leaf predicting:", stump1$surrogate$class2)
```

For predictors $(X_1,\dots,X_5)$, let $f$ denote the stump resulting from the best split and $\tilde{f}$ denote the stump resulting from the surrogate split. 

$$
f(X_1,\dots,X_5) = 
\begin{cases}
0 & X_1 = 0 \\
1 & X_1 = 1
\end{cases}\\
\tilde{f}(X_1,\dots,X_5) = 
\begin{cases}
0 & X_2 = 0 \\
1 & X_2 = 1
\end{cases}
$$ 

##### (a)(ii) Variable Importance 

Time to plug and chug. The best split is on $X_1$, and there is only one split in our stump, so we have only one value to compute, namely the importance for $X_1$. 

$$
\begin{aligned}
\text{Imp}^T(X_1) 
&= \Delta I(0,1,1) \\
&= \text{reduction in Gini Index at split} \\
&= 0.270
\end{aligned}
$$ 

To be honest, I don't know how to interpret this number. I'm not sure what it's range is or what is considered a large value. Relative to the surrogate split, which has importance 

$$
\text{Imp}^T(X_2) = 0.106,
$$

it seems that $X_1$ is more important to the classification. How to compare these values is not immediately obvious to me. One way I have seen to compare these values is to divide them all by the magnitude of the largest, in which case we would get a 'relative' importance like so. 

$$
\hat{Imp}^T(X_1) = 1, \quad \hat{Imp}^T(X_2) = 0.393
$$

With this measure, I feel more comfortable asserting that, yes, $X_1$ is more important than $X_2$, by a factor of about 2.5.

##### (a)(iii) MSE 

```{r testMSE}
library(dplyr)
test.MSE1 = test %>%
  mutate(yhat1 = ifelse(X1 == 0, 0, 1)) %>%
  mutate(error1 = (yhat1 - Y)^2) %>%
  mutate(yhat2 = ifelse(X2 == 0, 0, 1)) %>%
  mutate(error2 = (yhat2 - Y)^2)
MSE1 = mean(test.MSE1$error1)
MSE2 = mean(test.MSE1$error2)
knitr::kable(data.frame(MSE1, MSE2), col.names = c("MSE for X_1 split", "MSE for X_2 split"),
             caption = "Test MSEfor Best (X_1) and Surrogate (X_2) Split")
``` 

#### (b) Growing a Forest (of stumps)  

```{r growForest, cache=TRUE}
n = nrow(train)
B = 0.8 * n
M = 1000
K = 5
res = data.frame(leaf0 = rep(NA, M), split = rep(NA, M), leaf1 = rep(NA, M), surrogate = rep(NA, M),
                 delta = rep(NA, M), OOBerror = rep(NA, M), OOBperm = rep(NA, M))
forest = list(res,
              res,
              res,
              res,
              res)
for (k in 1:K){
  for (m in 1:M){
    bootstrap = sample(1:n, B, replace = TRUE)
    df.boot = train[bootstrap,]
    df.oob = train[-unique(bootstrap),]
    predictors = names(df.boot)[names(df.boot) != "Y"]
    df.boot = df.boot %>%
      select(sample(predictors, k), Y)
    stump = grow.stump(df = df.boot, surrogate = ifelse(k==1, FALSE, TRUE))
    forest[[k]]$leaf0[m] = stump$leaf1$class
    forest[[k]]$split[m] = stump$leaf1$split
    forest[[k]]$leaf1[m] = stump$leaf2$class
    if (k != 1){
      forest[[k]]$surrogate[m] = stump$surrogate$split
    }
    forest[[k]]$delta[m] = stump$deltaI
    df.oob = df.oob %>%
      mutate(yhat = ifelse(.[,stump$leaf1$split] == 0, 
                           stump$leaf1$class,
                           stump$leaf2$class))
    forest[[k]]$OOBerror[m] = sum(df.oob$yhat != df.oob$Y)/nrow(df.oob)
    df.oob[,stump$leaf1$split] = sample(df.oob[,stump$leaf1$split])
    df.oob = df.oob %>%
      mutate(yhat = ifelse(.[,stump$leaf1$split] == 0, 
                           stump$leaf1$class,
                           stump$leaf2$class))
    forest[[k]]$OOBperm[m] = sum(df.oob$yhat != df.oob$Y)/nrow(df.oob)
  }
}
```

Yes, the above would run faster if we did it in parallel, one core for each `k`, but it would take longer to write and debug that code than it does to run the code in its current state. User efficiency beats machine efficiency this time.  

##### (b)(i) Basic Summary of Forest 

Asking 'how many times was each variable the best split?' is not relevant when there are only $k=1$ choices for variables, so we omit that case here. A same triviality with with the surrogate splits for $k = 2$, but we do not omit this case.  

```{r forestTable, echo=FALSE}
bests = lapply(2:K, function(k){table(forest[[k]]$split)})
surrogates = lapply(2:K, function(k){table(forest[[k]]$surrogate)})
table.df = data.frame(matrix(NA, nrow = 4, ncol = 10))
rownames(table.df) = c("k=2", "k=3", "k=4", "k=5")
vars = c("X1", "X2", "X3", "X4", "X5")
for (i in 1:4){
  for (j in 1:5){
    table.df[i, j] = bests[[i]][vars[j]]
    table.df[i, (j+5)] = surrogates[[i]][vars[j]]
  }
}
colnames(table.df) = c("X1 best", "X2 best", "X3 best", "X4 best", "X5 best",
                       "X1 surrogate", "X2 surrogate", "X3 surrogate", "X4 surrogate", "X5 surrogate")
table.df[is.na(table.df)] = 0
knitr::kable(table.df, caption = "Occurences of Each Variable as Best or Surrogate Split")
```  


We see from the table that $X_1$ is overwhelmingly chosen to be the best variable for splitting; when all variables are available for splitting, $X_1$ appears in 100% of cases with $X_2$ always as its surrogate. This fact is significant considering that each tree is grown from a different bootstrap sample. The numbers we see here suggest that $X_1$ is the most important variable, followed by $X_2$.  

##### (b)(ii) Variable Importance by OOB Error

```{r oobTable, echo=FALSE}
table.df.imp = table.df.oob = data.frame(matrix(NA, nrow = 5, ncol = 5))
colnames(table.df.imp) = colnames(table.df.oob) = vars
for (i in 1:5){
  for (j in vars){
    of.interest = forest[[i]]$split == j
    table.df.imp[i,j] = (1/M)*sum(forest[[i]]$delta[of.interest])
    table.df.oob[i,j] = (1/M)*sum(forest[[i]]$OOBperm[of.interest],
                                  -forest[[i]]$OOBerror[of.interest])
  }
}
rownames(table.df.imp) = rownames(table.df.oob) = c("k=1","k=2","k=3","k=4","k=5")
knitr::kable(table.df.imp, caption = "Variable Importance by Sum of Importance Across Trees", digits = 4)
knitr::kable(table.df.oob, caption = "Variable Importance by Out-of-Bag Error", digits = 4)
```  


Based on the tables we see above, $X_1$ appears to be the most important variable in every case, followed by $X_2$ in every case. 

Intuitively, the impact of masking should be lower in a forest. Each tree is grown on a random subset of predictors, so a dominant predictor cannot mask the others in every tree. It will not even be an option for splitting in some trees. Therefore, the other predictors will be used for splits and their importance appears within the importance measure for the forest as whole.  

##### (b)(iii) MSE in the Forest  

```{r forestMSE, cache=TRUE}
forest.vote = function(k, x.df){
  forest = forest[[k]]
  yhat = rep(0, nrow(x.df))
  for (i in 1:nrow(forest)){
    for (j in 1:nrow(x.df)){
      var.of.interest = forest$split[i]
      yhat.vote = ifelse(x.df[j,var.of.interest]==0,
                         forest$leaf0[i],
                         forest$leaf1[i])
      yhat[j] = yhat[j] + (yhat.vote - .5)
    }
  }
  final_vote = 1*(yhat > 0)
  return(final_vote)
}
trees.ind.class = function(k, test.df){
  Y = test.df$Y
  forest = forest[[k]]
  loss = rep(NA, nrow(forest))
  for (i in 1:nrow(forest)){
    error = rep(NA, nrow(test.df))
    for (j in 1:nrow(test.df)){
      var.of.interest = forest$split[i]
      yhat = ifelse(test.df[j,var.of.interest]==0,
                         forest$leaf0[i],
                         forest$leaf1[i])
      error[j] = (Y[j] - yhat)^2
    }
    loss[i] = mean(error)
  }
  return(mean(loss))
}
test.x = test[1:(ncol(test) - 1)]
test.df = test %>%
  mutate(yhat.vote1 = forest.vote(k = 1, test.x)) %>%
  mutate(yhat.vote2 = forest.vote(k = 2, test.x)) %>%
  mutate(yhat.vote3 = forest.vote(k = 3, test.x)) %>%
  mutate(yhat.vote4 = forest.vote(k = 4, test.x)) %>%
  mutate(yhat.vote5 = forest.vote(k = 5, test.x)) %>%
  mutate(error1 = (Y-yhat.vote1)^2) %>%
  mutate(error2 = (Y-yhat.vote2)^2) %>%
  mutate(error3 = (Y-yhat.vote3)^2) %>%
  mutate(error4 = (Y-yhat.vote4)^2) %>%
  mutate(error5 = (Y-yhat.vote5)^2) 
MSE.forest = data.frame(vote = rep(NA, 5), ind = rep(NA, 5))
for (i in 1:5){
  MSE.forest[i, "vote"] = mean(test.df[,(i+11)])
  MSE.forest[i, "ind"] = trees.ind.class(k = i, test.df = test)
}
rownames(MSE.forest) = rownames(table.df.imp)
knitr::kable(MSE.forest, col.names = c("Loss by Vote", "The Weird Loss"),
             caption = "Comparison of Loss Measures")
```  


The voting method is the proper method for calculating loss since that's how random forests actually make classifications of new data. We want our estimate of loss to come from the process actually used in classification. 